---
title: "Reproducible Research Assignament 1"
author: "Juan M. López"
date: "Thursday, May 14, 2015"
output: html_document
---

##Loading and preprocessing the data

```{r echo = TRUE}
data <- read.csv("activity.csv", stringsAsFactors = FALSE)
```

##What is mean total number of steps taken per day?

###Number of steps per day
```{r echo=TRUE}
tapply(data$steps[!is.na(data$steps)], data$date[!is.na(data$steps)],sum)
```

###Histogram
```{r echo=TRUE}
steps_day <- tapply(data$steps[!is.na(data$steps)], data$date[!is.na(data$steps)],sum)
hist(steps_day,xlab="Steps",ylab="Days", main="Steps per Day")

```

### Mean and median of the total number of steps taken per day
```{r echo=TRUE}
print(paste("Mean: ", mean(steps_day[!is.na(steps_day)])))
print(paste("Median: ", median(steps_day[!is.na(steps_day)])))

```

##What is the average daily activity pattern?

###Plot: Interval vs Average Steps
```{r echo=TRUE}
avg_steps <- tapply(data$steps[!is.na(data$steps)], data$interval[!is.na(data$steps)],mean)
with(data, plot(interval[1:288], avg_steps, type="l", xlab="Interval", ylab="Average Steps"))

```

###Interval with Max Steps
```{r echo=TRUE}
print(paste("The Interval with the maximum number of steps is the", names(avg_steps[match(max(avg_steps), avg_steps)]), "th with an average of", max(avg_steps), "steps."))

```

##Imputing missing values

###Calculate and report the total number of missing values
```{r echo=TRUE}
print(paste("The total number of missing values is:", sum(is.na(data))))
```

###Create a new dataset that is equal to the original dataset but with the missing data filled in

I will use the mean of the 5-min interval to fill the missing values
```{r echo=TRUE}
na_index <- which(is.na(data$steps)==TRUE)

for(i in na_index){
        data$steps[i] <- avg_steps[match(data$interval[i],names(avg_steps))]                
}
```

###Make a histogram of the total number of steps taken each day and Calculate and report the mean and median

I will use the mean of the 5-min interval to fill the missing values
```{r echo=TRUE}
steps_day <- tapply(data$steps, data$date,sum)
hist(steps_day,xlab="Steps",ylab="Days", main="Steps per Day")

print(paste("Mean: ", mean(steps_day)))
print(paste("Median: ", median(steps_day)))
```

As you may see, with this method the mean and the median show almost no change (less than 1%), in fact, watching the histogram, we can see the higher values are even emphasized.  

##Are there differences in activity patterns between weekdays and weekends?

###Create a new factor variable in the dataset with two levels - "weekday" and "weekend"
```{r echo=TRUE}
library(dplyr)
library(lubridate)

data$date <- strptime(data$date, "%Y-%m-%d")
data <- mutate(data, day = ifelse(wday(date) <= 5,"weekday","weekend"))
data$day <- as.factor(data$day)

```

###Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)

```{r echo=TRUE}
data <- split(data,data$day)

avg_steps_1 <- tapply(data$weekday$steps, data$weekday$interval,mean)
avg_steps_2 <- tapply(data$weekend$steps, data$weekend$interval,mean)

par(mfrow = c(1,2), pin=c(2,2))
with(data$weekday, plot(interval[1:288], avg_steps_1, type="l", xlab="Interval", ylab="Average Steps", main="Weekday"))
with(data$weekend, plot(interval[1:288], avg_steps_2, type="l", xlab="Interval", ylab="Average Steps", main="Weekend"))

```
